# (PART) Clinical Trial Project {-}

# Overview

In a late-stage clinical trial, the number of A&R deliverables can easily be in the hundreds. 
For an organization, it is also common to have multiple ongoing clinical trials in a clinical program. 

To deliver A&R results of a clincial tiral project, it is a team work that typically require
collaborations from both statisticians and programmers. 
In this part, let's consider how to organize a clinical trial project as an A&R leader. 

In Chapter \@ref(folder), we will discuss how to organize source code, documents, 
deliverables in an A&R clinical project. 
We recommend to use the R package folder structure. 

In Chapter \@ref(manage), we will discuss a process or system development lifecycle 
to manage A&R of a clinical project. 
We recommend to follow an agile management approach to define, develop, validate and deliver work.

# Project Folder {#folder}

A clearly defined clinical project folder structure can have many benefits to 
clinical development teams in an organization. 
Specifically a well defined project folder can achieve:

- Consistency: every team work on the same folder structure.
- Reproducibility: analysis can be executed and reproduced by different team members months/years later.  
- Automation: automatically check integration of a project.  
- Compliance: reduce compliance issue

We illustrate the R package project folder for A&R project using a demo project called 
[`esubdemo`](https://github.com/elong0527/esubdemo.git). You can [clone](https://happygitwithr.com/rstudio-git-github.html#clone-the-new-github-repository-to-your-computer-via-rstudio) 
the project using RStudio. 

## Consistency 

For consistency, a well defined folder structure with potential templates
ensures project teams organize A&R work consistently across multiple projects. 
Consistent folder structure also reduce communication cost between study team members 
and enhances the transparency of projects.

For R user, you already benefit from a well defined and consistent folder structure. 
That is the [R package folder structure](https://github.com/rstudio/cheatsheets/raw/master/package-development.pdf). 
Every R package developer is required to follow the same convention to organize their R
functions, before the R package can be disseminated through the Comprehensive R Archive Network
(CRAN). As a user, you can easily install and use those R packages after installing from CRAN. 
There are many good resources to guide developers on R package development.
For example, [Hadley Wickmans' R packages book](https://r-pkgs.org/)

We recommend to use R package folder structure to organize analysis scripts 
for clinical trial development. 
Using the R package folder structure to streamline data analysis work has also been proposed before 
(@marwick2018packaging, @wuanalysis). 

In this book, we refer an R package as **project specific R package**. 
The purpose of project specific R package is to organize analysis scripts for a clinical project.
We refer an R package as **standard R package** if the purpose of an R package is to 
share commonly used R functions to be hosted in an code repository such as CRAN. 

Below is a minimal sufficient folders and files based on the R package folder structure.  

- `*.Rproj`: RStudio project file used to open RStudio project.
- `DESCRIPTION`: Metadata for a package including authors, license, dependency etc.
- `R/`: Project specific R functions.
- `vignettes/`: Analysis scripts using Rmarkdown.
- `man/`: Manual of project specific R functions.

To illustrate the idea, you can open `esubdemo` project by click the `esubdemo.Rproj` project. 

For analysis code to create TLFs, we saved them in the `vignettes/` folder. 
In `esubdemo`, we put analysis code we learned from previous chapters into the `vignettes` folder.

> Here the `adam\` folder contain ADaM datasets. 
> The `tlf` folder contain output TLFs in `rtf` and `pdf` format.
> We put `adam\` and `tlf\` folder within `vignettes` folder only for illustration purpose. 
> In a real A&R report, you may have different location to save your input and output.

```
vignettes
├── adam
├── tlf
├── tlf-01-disposition.Rmd
├── tlf-02-population.Rmd
├── tlf-03-baseline.Rmd
├── tlf-04-efficacy.Rmd
├── tlf-05-ae-summary.Rmd
└── tlf-06-ae-spec.Rmd
```

In previous chapters, we also create a few helper functions that can be used for different analysis. 
(e.g. `fmt_num` and `count_by`). 
We put those helper functions into the `R/` folder. 

> Here `zzz.R` is a template file that ensure startup file `inst/startup.R` is executed. 
> We will revisit startup file while we discuss reproducibility. 

```
R
├── count_by.R
├── fmt.R
└── zzz.R
```

The `man` folder is used to save documentation automatically generated by `roxygen2`.
A typical workflow is to add `roxygen2` documentations at the header of each function in the `R\` folder. 
Then `devtools::documents()` is used to generate all the documentation files in the `man\` folder. 
More details can be found in [Chapter 10 of the R packages book](https://r-pkgs.org/man.html). 

## Reproducibility {#reproduce}

Reproducibility of analysis can be challenge. 
By using R package folder structure and proper tools (e.g. `renv`, `packrat`), 
we can easily achieve reproducibility for R and R package versions. 

> This is the same level of reproducibility in most SAS environment.
> https://support.sas.com/techsup/pcn/altopsys.html

In `esubdemo` project, a reproducible environment is created when you open the 
`esubdemo.Rproj` from RStudio. 
While RStudio open `esubdemo` project, RStudio will automatically execute R code in `.Rprofile`.
So we are using `.Rprofile` to setup a reproducible environment. 
More details can be found in [https://rstats.wtf/r-startup.html](https://rstats.wtf/r-startup.html). 

> `.Rprofile` is only for project specific R package. 
> A standard R package should not use `.Rprofile`.   

After the `esubdemo` project is open, the code in `.Rprofile` will automatically check 
current R version is the same as we defined in `.Rprofile` 

```{r, eval = FALSE}
R_version <- "4.0.1"	# set up project R version
```

If the R version mismatch, an error message is displayed as below.

```
Error: The current R version is not the same as the current project in R4.0.1
```

If we are running under the same R version, 
you shall be able to see similar messages to inform how we control R package version. 

```
* Project '~/esubdemo' loaded. [renv 0.14.0]
Current oroject R package repository:
    https://mran.microsoft.com/snapshot/2021-08-06
```

The R package version is controlled in two layers. 
Firstly, we define a snapshot date in `.Rprofile`. 
The snapshot date allows to freeze the repository. 
In other words, all R packages installed in this RStudio project 
are based on the frozen R version at the snapshot date. 
Here it is `2021-08-06` by using Microsoft R Application Network (MRAN) server. 

```{r, eval = FALSE}
snapshot  <- "2021-08-06" 									                           # set up snapshot date
repos     <- paste0("https://mran.microsoft.com/snapshot/", snapshot)  # set up repository based on snapshot
```

> RStudio Package Manager (RSPM) provided a solution to host both 
> public avaiable and internally developed R packages. 
> However, public RSPM server (https://packagemanager.rstudio.com/client/#/repos/2/overview) 
> did not provide daily snapshot as MRAN had. 

Secondly, we use `renv` to lock R package versions and saved in `renv.lock` file. 
The exact same R package versions can be restored with `renv::restore()`.
An introduction of `renv` can be found in its [package website](https://rstudio.github.io/renv/articles/renv.html).

```
* Project '~/github-repo/esubdemo' loaded. [renv 0.14.0]
```

> `renv.lock` file and `renv/` folder are only for project specific R package. 
> A standard R package should not use `renv`.   

Once R packages have been properly installed, 
the system will use the R packages located in the search path defined based on the order of `.libPaths()`.
The startup message also provided the R package search path. 

```
Below R package path are searching in order to find installed R pacakges in this R session:
    /home/esubdemo/renv/library/R-4.0/x86_64-pc-linux-gnu
    /rtmp/Rtmp8d5mcL/renv-system-library
```

For an organization, a cloud based R environment (e.g. RStudio Workbench) 
can enhance the development and operation of a clinical development team by 
centralize managing R and R package version. 
More details can be found in [https://environments.rstudio.com/](https://environments.rstudio.com/).   

In summary, to achieve reproducibility for a project specific R package, 
a clinical project team can work under a controlled R environment 
in the same R version and R package versions defined by a repository snapshot date.

## Automation 

By using the R package folder structure, you can also benefit from many outstanding tools to 
simplify and streamline your workflow. 

There are additional folders and files required in this demo for an analysis project

> people may use different folder name as it is not a standard R package folder.
> These folders and files needs to be added in `.Rbuildignore` to pass
> R package check 

- `adam`: ADaM datasets in `.sas7bdat` format.
  + one may also put it in `inst/exdata` following [R package convention](https://r-pkgs.org/data.html)
  + in reality, we suggest to have real data saved outside of this project. (e.g. in a database)
- `output`: TLFs output 
- `renv.lock` and `renv`: R package management using `renv` package. ([Introduction](https://rstudio.github.io/renv/articles/renv.html)) 
- `_pkgdown.yml`: [pkgdown](https://pkgdown.r-lib.org/articles/pkgdown.html) configuration file
- `.Rprofile`: Project startup file to setup running environment including R version, repository, folder path etc. 
  - We further use `inst/startup.R` and `R/zzz.R` to allow the startup file is executed while running. `devtools::load_all()` and RStudio build panel. 
  
  
## Compliance

- ToDo
