# Disposition {#disposition}

Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
we need to summarize accounting of all patients who entered the study, 
in the section 10.1, disposition of patients of a CSR. 

```{r}
library(haven) # Read SAS data
library(dplyr) # Manipulate data
library(tidyr) # Manipulate data
library(r2rtf) # Reporting in RTF format
```

In this chapter, we illustrate how to create a disposition table as below.

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("tlf/tbl_disp.pdf")
```

The first step is to read relevant datasets into R. 
For disposition table, all the required information is saved in the ADSL dataset.
We can use `haven` package to read the dataset.

```{r}
adsl <- read_sas("adam_data/adsl.sas7bdat")
```

We illustrate how to prepare a report data for a simplified disposition 
of patients table using variables below: 

- USUBJID: Unique subject identifier 
- TRT01P: Planed treatment
- TRT01PN: Planned treatment encoding
- DISCONFL: Discontinued from study flag
- DCREASCD: Discontinued from study reason 

```{r}
adsl %>%
  select(USUBJID, TRT01P, TRT01PN, DISCONFL, DCREASCD) %>%
  head(4)
```
In the code below, we calculated patients in the analysis population. 

```{r}
n_rand <- adsl %>%
  group_by(TRT01PN) %>%
  summarize(n = n()) %>%
  pivot_wider(
    names_from = TRT01PN,
    names_prefix = "n_",
    values_from = n
  ) %>%
  mutate(row = "Patients in population")
n_rand
```
In the code below, we calculate number and percentage of patients who discontinued the study. 

Here we use `formatC()` to customize the numeric value to be 1 decimal with fixed width of 5. 

```{r}
n_disc <- adsl %>%
  group_by(TRT01PN) %>%
  summarize(
    n = sum(DISCONFL == "Y"),
    pct = formatC(n / n() * 100,
      digits = 1, format = "f", width = 5
    )
  ) %>%
  pivot_wider(
    names_from = TRT01PN,
    values_from = c(n, pct)
  ) %>%
  mutate(row = "Discontinued")

n_disc
```
In the code below, we calculate number and percentage of patients 
who completed/discontinued the study in different reasons. 

```{r}
n_reason <- adsl %>%
  group_by(TRT01PN) %>%
  mutate(n_total = n()) %>%
  group_by(TRT01PN, DCREASCD) %>%
  summarize(
    n = n(),
    pct = formatC(n / unique(n_total) * 100,
      digits = 1, format = "f", width = 5
    )
  ) %>%
  pivot_wider(
    id_cols = DCREASCD,
    names_from = TRT01PN,
    values_from = c(n, pct),
    values_fill = list(n = 0, pct = "  0.0")
  ) %>%
  rename(row = DCREASCD)

n_reason
```

In the code below, we calculate number and percentage of patients 
who complete the study. We split `n_reason` because we need to customize 
the row order of the table.

```{r}
n_complete <- n_reason %>% filter(row == "Completed")

n_complete
```
In the code below, we calculate number and percentage of patients 
who discontinued the study in different reasons. 
Here we use `paste0("    ", row)` to add some leading space for individual reasons for study discontinuation. 

```{r}
n_reason <- n_reason %>%
  filter(row != "Completed") %>%
  mutate(row = paste0("    ", row))

n_reason
```

Now we combined individual rows into the whole table for reporting purpose. 
`tbl_disp` is used as input for `r2rtf` to create final report. 

```{r}
tbl_disp <- bind_rows(n_rand, n_complete, n_disc, n_reason) %>%
  select(row, ends_with(c("_0", "_54", "_81")))

tbl_disp
```

We start to define the format of the output. We highlighted items that 
is not discussed in previous discussion.

`rtf_title` define table title. We can provide a vector for the title argument. 
Each value is a separate line. 
The format can also be controlled by providing a vector input in text format. 


```{r}
tbl_disp %>%
  
  # Table title
  rtf_title("Disposition of Patients") %>%
  
  # First row of column header
  rtf_colheader(" | Placebo | Xanomeline Low Dose| Xanomeline High Dose",
    col_rel_width = c(3, rep(2, 3))
  ) %>%
  
  # Second row of column header
  rtf_colheader(" | n | (%) | n | (%) | n | (%)",
    col_rel_width = c(3, rep(c(0.7, 1.3), 3)),
    border_top = c("", rep("single", 6)),
    border_left = c("single", rep(c("single", ""), 3))
  ) %>%
  
  # Table body
  rtf_body(
    col_rel_width = c(3, rep(c(0.7, 1.3), 3)),
    text_justification = c("l", rep("c", 6)),
    border_left = c("single", rep(c("single", ""), 3))
  ) %>%
  
  # Encoding RTF syntax
  rtf_encode() %>%
  
  # Save to a file
  write_rtf("tlf/tbl_disp.rtf")
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("tlf/tbl_disp.pdf")
```
