# Disposition {#disposition}

Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf), a summary needs to be provided that accounts for all participants who entered the study within Section 10.1, Disposition of participants.
This section summarizes the number of participants that were randomized, entered and completed each phase of the study. In addition, the reasons for all post-randomization discontinuations, grouped by treatment and by major reason of discontinuation are required (lost to follow-up, adverse event, ...).

```{r}
library(haven) # Read SAS data
library(dplyr) # Manipulate data
library(tidyr) # Manipulate data
library(r2rtf) # Reporting in RTF format
```

In this chapter, the creation of a typical disposition table is shown.

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("tlf/tbl_disp.pdf")
```

The first step is to read in the relevant datasets into R.
For a disposition table, all the required information is saved in a Subject-level Analysis Dataset (ADSL).
The dataset is provided in SAS7bdat format, which is the current standard in current clinical trial reporting and analysis.The `haven` package is able to read in the dataset, while maintaining its attributes (eg. variable labels).

```{r}
adsl <- haven::read_sas("adam_data/adsl.sas7bdat")
```

Following variables are used in the preparation of the disposition report:

-   USUBJID: Unique subject identifier
-   TRT01P: Planed treatment
-   TRT01PN: Planned treatment numeric encoding
-   DISCONFL: Discontinued from study flag
-   DCREASCD: Discontinued from Study reason coded

```{r}
adsl %>%
  dplyr::select(USUBJID, TRT01P, TRT01PN, DISCONFL, DCREASCD) %>%
  utils::head(4)
```

In the code below, the number of participants randomized is calculated.

```{r}
n_rand <- adsl %>%
  dplyr::group_by(TRT01P) %>%
  dplyr::summarize(n = n()) %>%
  tidyr::pivot_wider(
    names_from = TRT01P,
    names_prefix = "",
    values_from = n
  ) %>%
  dplyr::mutate(row = "participants Randomized")

n_rand
```

In the code below, the number of participants that discontinued the study is calculated.

Here we use the `formatC()` function to customize the numeric value to be 1 decimal with fixed width of 5.

```{r}
n_disc <- adsl %>%
  dplyr::group_by(TRT01P) %>%
  dplyr::summarize(
    n = base::sum(DISCONFL == "Y"),
    pct = base::formatC(n / n() * 100,
      digits = 1, format = "f", width = 5
    )
  ) %>%
  dplyr:: mutate(npct = base::paste0(n, " (", pct, "%)")) %>%
  dplyr::select(-c(n,pct)) %>%
  tidyr::pivot_wider(
    names_from = TRT01P,
    values_from = npct,
    values_fill = list(npct = "0")
  ) %>%
  dplyr::mutate(row = "Discontinued")

n_disc
```

To calculate the number and percentage of participants that completed/discontinued the study for each discontinuation reason, below code can be used.

```{r}
n_reason <- adsl %>%
  dplyr::group_by(TRT01P) %>%
  dplyr::mutate(n_total = dplyr::n()) %>%
  dplyr::group_by(TRT01P, DCREASCD) %>%
  dplyr::summarize(
    n = dplyr::n(),
    pct = base::formatC(n / unique(n_total) * 100,
      digits = 1, format = "f", width = 5
    )
  ) %>%
  dplyr:: mutate(npct = base::paste0(n, " (", pct, "%)")) %>%
  tidyr::pivot_wider(
    id_cols = DCREASCD,
    names_from = TRT01P,
    values_from = npct,
    values_fill = list(npct = "0")
  ) %>%
  dplyr::rename(row = DCREASCD)

n_reason
```

From the step above, the number and percentage of participants that completed the study can be isolated so change the
roworder in the final table.

```{r}
n_complete <- n_reason %>% 
  dplyr::filter(row == "Completed")

n_complete
```

For display purposes leading spaces are added to study discontinuation. This will mimick identation in the final report.

```{r}
n_reason <- n_reason %>%
  dplyr::filter(row != "Completed") %>%
  dplyr::mutate(row = paste0("    ", row))

n_reason
```

In the final step, the individual rows are combined and used as input for `r2rtf.

```{r}
tbl_disp <- base::rbind(n_rand, n_complete, n_disc, n_reason)

tbl_disp
```

In the below code, formatting of the final table is defined. Sections are highlighted that were not discussed before.

```{r}
tbl_disp %>%
  
  # Table title
  rtf_title("Disposition of participants") %>%
  
  # First row of column header
  rtf_colheader(" | Placebo | Xanomeline Low Dose| Xanomeline High Dose",
    col_rel_width = c(3, rep(2, 3))
  ) %>%
  
  # Second row of column header
  rtf_colheader(" | n | (%) | n | (%) | n | (%)",
    col_rel_width = c(3, rep(c(0.7, 1.3), 3)),
    border_top = c("", rep("single", 6)),
    border_left = c("single", rep(c("single", ""), 3))
  ) %>%
  
  # Table body
  rtf_body(
    col_rel_width = c(3, rep(c(0.7, 1.3), 3)),
    text_justification = c("l", rep("c", 6)),
    border_left = c("single", rep(c("single", ""), 3))
  ) %>%
  
  # Encoding RTF syntax
  rtf_encode() %>%
  
  # Save to a file
  write_rtf("tlf/tbl_disp.rtf")
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("tlf/tbl_disp.pdf")
```

The procedure to generate a disposition table can be summarized as follow:

-   Step 1: Read subject level data into R, i.e. `adsl`.
-   Step 2: Count participants in the analysis population and name the dataset `n_rand`.
-   Step 3: Count the number and percentage of participants that discontinued the study, and name the dataset `n_disc`.
-   Step 4: Count the number and percentage of participants that discontinued the study for different reasons, and name the dataset `n_reason`.
-   Step 5: Count the number and percentage of participants that completed the study, and name the dataset `n_complete`.
-   Step 6: Bind `n_rand`, `n_disc`, `n_reason` and `n_complete` by row. 
-   Step 7: Write the final table to RTF
